---
title: "Assessing model fit"
author: "Amy Willis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assessing model fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## The importance of verifying assumptions

As with all statistical models, the assumptions of `DivNet` should be verified before interpreting the results. In this vignette, I will show you how to investigate a key assumption of the `DivNet` model: that the log-ratio relative abundances are linear (or at least not strongly curved).

You do not need to do this if you are fitting `DivNet` with categorical covariates, only if you are fitting `DivNet` with continuous covariates.

## Testing the changes in Shannon diversity with temperature

We are going to use the Lee et al dataset and investigate the effect of temperature on phylum-level Shannon diversity. Recall that this dataset contains samples of microbial communities living on seafloor rocks near a hydrothermal vent.

We first load the data and necessary packages, aggregate counts to the phylum level, and subset to samples with temperature data.

```{r}
if ("speedyseq" %in% row.names(installed.packages())) { 
  library(speedyseq) # if speedyseq is installed, use it for tax_glom 
} else {
  library(phyloseq) # if not, use phyloseq for tax_glom
}
library(breakaway)
library(DivNet)
library(magrittr)
library(ggplot2)
data(Lee)
lee_phylum <- Lee %>%
  tax_glom(taxrank = "Phylum") %>%
  subset_samples(temp != "<NA>")
```

Annoyingly, temperature is a categorical variable in this dataset, so we convert it to a continuous variable and run `DivNet`, modelling the log-ratios as a linear function of temperature:

```{r, message=FALSE}
lee_phylum %>% sample_data %$% temp
temp_continuous <- lee_phylum %>% sample_data %$% temp %>% as.character %>% as.numeric
divnet_temperature <- lee_phylum %>%
  divnet(X = temp_continuous, base=1)
```

I'm a big fan of plotting the estimated diversity against the covariate before doing any hypothesis testing to make sure that everything looks reasonable.

```{r}
shannon_table <- divnet_temperature %$%
  shannon %>%
  summary %>%
  cbind(temp_continuous)
shannon_table %>%
  ggplot(aes(x = temp_continuous, y = estimate)) + geom_point() +
  geom_linerange(aes(x = temp_continuous, ymin = lower, ymax = upper))
```

Looks good to me! Let's go ahead and see if that upward trend is statistically significant given those wide error bars:

```{r}
betta(shannon_table$estimate, shannon_table$error,
      cbind(1, divnet_temperature$X))$table
```

We reject the null hypothesis of no significant linear increase in Shannon diversity with temperature with $p < 10^{-3}$.

# Assessing model fit

Before taking that result too seriously, let's check for model misspecification. The `DivNet` model performs well when there is not significant curvature in the log-ratios. To investigate this, let's pull out the observed log-ratios (`observed_yy`) as well as the fitted log-ratios (`fitted_yy`).

```{r}
fitted_yy <- DivNet::to_log_ratios(divnet_temperature$fitted_z,
                                   base = 1)
observed_yy <- lee_phylum %>% otu_table %>%
  as.matrix %>% t %>%
  DivNet::to_log_ratios(base = 1)
```

Now let's put both the observed log-ratios and the fitted log-ratios against temperature. `DivNet` assumes a linear model for all log-ratios, and since there are 20 taxa in this dataset, there are 19 log-ratios. Therefore I'm going to write a function that plots the log-ratios.

```{r}
draw_log_ratios <- function(taxon) {
  data.frame(temperature = temp_continuous,
             log_ratio = fitted_yy[,taxon],
             log_ratio_observed = observed_yy[,taxon]) %>%
    ggplot(aes(x = temperature, y = log_ratio_observed)) +
    geom_point() +
    xlab("Temperature") +
    ylab("log-ratios") +
    geom_point(aes(y = log_ratio), col = "blue") +
    ggtitle("Observed (black) and fitted (blue) log-ratios against temperature")
}
draw_log_ratios(taxon=1)
draw_log_ratios(taxon=2)
draw_log_ratios(taxon=3)
```

I don't see any significant curvature in any of these plots. Note that it's not important that these scatterplots display a strong linear trend -- just that they are not strongly nonlinear. Note that we found that model is robust to mild nonlinearity, but its performance decreases as the nonlinear trend gets stronger. 

In theory I should run the function and inspect those plots, but given that there may be hundreds or thousands of taxa model, I would recommend checking the most abundant taxa, since these are the ones that are most likely to throw off diversity estimation if misspecified.

Happy modelling!
