---
title: "Understanding DivNet"
noauthor: "Amy Willis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding DivNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Amy tries to unravel some curious behaviour in DivNet

```{r}
devtools::install_github("adw96/breakaway")
library(breakaway)
library(magrittr)
library(phyloseq)
library(DivNet)
data(Lee)
Lee
lee_phylum <- tax_glom(Lee, taxrank="Phylum")
```

```{r, results = "hide"}
divnet_phylum_char <- lee_phylum %>%
  divnet(X = "char", ncores = 4)
```

Let's now compare the plug-in Shannon index with the divnet estimates

```{r}
library(ggplot2)
dv_plot <- divnet_phylum_char$shannon %>% 
  plot(lee_phylum, color = "char") +
  xlab("Basalt characteristic") +
  ylab("Shannon diversity estimate\n(phylum level)") +
  coord_cartesian(ylim = c(0,2))

naive_plot <- lee_phylum %>%
  sample_shannon %>%
  plot(lee_phylum, color = "char")  +
  coord_cartesian(ylim = c(0,2))

ggpubr::ggarrange(dv_plot, naive_plot)
```

## Why are the altered estimates so much lower than the sample estimates?

Let's try subsetting into just altered and carbonate:

```{r}
dv_a <- lee_phylum %>%
  subset_samples(char %in% c("altered", "carbonate")) %>% divnet(X = "char", ncores = 4)
dv_a$shannon %>% 
  plot(lee_phylum, "char") +
  coord_cartesian(ylim = c(0,2))
```

The estimates are about 1.4, which is what we'd expect. Let's try other combinations

```{r}
dv_ab <- lee_phylum %>%
  subset_samples(char %in% c("altered", "biofilm")) %>% 
  divnet(X = "char", ncores = 4)
dv_ab$shannon %>% 
  plot(lee_phylum, "char") +
  coord_cartesian(ylim = c(0,2))
```

Now they are lower! Hmmm. 

```{r}
lee_phylum %>%
  sample_richness %>%
  plot(lee_phylum, color = "char") 
```

Ok, so the altered specimens have much higher richness than the other types. I wonder if that's leading to the discrepancy?

```{r}
lee_phylum %>% otu_table
lee_phylum %>% sample_data
lee_phylum %>% 
  plot_bar()
```